\section{AMQP for Jolie}
In this chapter we will describe our implementation of the extension Advanced Message Queue Protocol for Jolie and we will argue for the design choices we have made. We use a RabbitMQ\cite{RabbitMQ} server as our message queue and exchange server.

\subsection{Jolie integration}
We are writing an extension to Jolie, and for it to be useful, it will have to work as an integration and work seamlessly with Jolie.

\subsubsection{Interfaces}
Jolie has a number of interfaces for its extensions and protocols to implement. The interfaces we are going to implement are:
\begin{itemize}
\item CommChannel
\item CommChannelFactory
\item CommListener
\item CommListenerFactory
\end{itemize}

\#\#\# Michael verify the following and maybe extend.

CommChannels are used to handle CommMessages for both output ports and input ports.\\
CommListeners are used to wait for requests on input ports. When a request is received Jolie will create a CommChannel to handle the request.\\
The two factories are used by Jolie for creating CommChannels and CommListeners.

\subsubsection{Registering the extension}
After implementation, the extension is then registered in a manifest-file. The manifest files for CommChannel and CommListener extensions for Jolie needs to have the following two lines:
\begin{lstlisting}
  X-JOLIE-ChannelExtension: amqp:jolie.net.AmqpCommChannelFactory
  X-JOLIE-ListenerExtension: amqp:jolie.net.AmqpListenerFactory
\end{lstlisting}

The first line defines that for output ports that has a location starting with `amqp://' Jolie should use the `jolie.net.AmqpCommChannelFactory' class.
The second line defines that for input ports that has a location starting with `amqp://' Jolie should use the `jolie.net.AmqpListenerFactory' class.

\subsubsection{The CommChannel}
To be able to integrate AMQP with Jolie we need to understand how Jolie sends a request, and how the response is handled.

When Jolie tries to invoke a method on a service, it starts by creating an instance of the relevant CommChannel. To create this instance Jolie will use a factory that is a subtype of CommChannelFactory. In our case we would need to create two classes called AmqpCommChannel and AmqpCommChannelFactory. The CommChannelFactory class defines one method that must be overridden in subclasses; createChannel. This method is responsible for creating an instance of a subclass of the CommChannel class.

A CommChannel subclass is responsible for sending and receiving CommMessage objects. CommChannel defines the following methods that must be overridden:
\begin{itemize}
  \item public CommMessage recvResponseFor(CommMessage request)
  \item protected CommMessage recvImpl()
  \item protected void sendImpl(CommMessage message)
  \item protected void closeImpl()
\end{itemize}

We are not interested in implementing the recvResponseFor method, we implement another subclass of the CommChannel, the StreamingCommChannel (this implements AbstractCommChannel, which in turn implements CommChannel).

The sendImpl method is called whenever the Jolie-code is trying to send a message to another service. It receives the CommMessage object to send. It is the CommChannels responsibility to make sure that the protocol selected from the Jolie-code is used. The sendImpl should call the relevant protocols send method. This encodes the CommMessage object into an output stream. This output stream should then be sent to the service, and our subclass (AmqpCommChannel) should handle this for AMQP.

When something is sent with success the recvImpl method is called. It is important to note, that this method is called regardless of the type of service call. Even if the called method is of the OneWay type, recvImpl is still called, but with an empty return message. Again, it is the CommChannels responsibility to make sure that the protocol is used, and the recvImpl should call the relevant protocols recv method. This converts the input stream from the response into a CommMessage object.

When the Jolie program terminates the connections must be closed. The closeImpl method is called in this case, and our subclass must handle closing the AMQP connections.

\subsubsection{The CommListener}
To setup Jolie to listen for requests from other services, it used the CommListener object. To support handling requests from a type of location, one needs to create a subclass of the CommListener. Like the CommChannel, a factory class is also used. The CommListenerFactory which needs to be derrived. In our case we would have to create the classes called AmqpCommListener and AmqpCommListenerFactory. When Jolie starts a program that has an input port with an amqp location it invokes the createListener method of the factory for creating a listener.

The abstract class CommListener specifies the following methods that must be overridden:
\begin{itemize}
  \item public void run()
  \item public void shutdown()
\end{itemize}

The CommListener class implements the Java Thread class (actually, it implements JolieThread which in turn implements Thread), and when Jolie starts the thread the run method is called. This method should use whatever blocking call the location has for accepting a connection, and handle it using another thread to be able to immediately be ready for another connection.

The shutdown method is used when the listener is closed, and should handle closing connections.

\subsection{Data formats}
\label{subsec:Data formats}
We looked at multiple "AMQP clients" for Java and frameworks building upon these clients. The Advanced Message Queue protocol is, as its name suggest, not simple so we thought it smartest to reuse existing software and then writing our own layer between the reused client and Jolie. The most promising, most used and most updated seemed to be the basic AMQP client build by RabbitMQ\cite{RabbitMqClient}. RabbitMQ's server and client is widely used by a number of large companies and is also sold as part of a VMWare software package\cite{vFabric}. The client is freely available from RabbitMQ's website.

The RabbitMQ client handles a lot of the AMQP stuff such as establishing connections and channels. The client also provides methods for publishing and subscribing. The publish method takes the body of the message as a byte-array. We have chosen to use this publish-method to send messages. That means converting everything we send to byte-arrays and thereby we do not use the AMQP described data types.

When everything is handled as byte arrays, we do not yet need to worry about the types of the AMQP specification. To fully support communication with other AMQP services, we would need to implement the type-system of AMQP, but we chose to skip this part, to be able to focus on other, more important aspects of the implementation.

For more about the AMQP Type System, see the protocol specification\cite{AmqpTypes}.

\subsection{OneWay}
The first part of our implementation is about publishing messages to an exchange on our AMQP server. This meant implementing the OneWay type for methods on services. This section will describe our implementation of the OneWay method type.

To be able to publish messages to an exchange (which in turn routes it to a queue) one must specify the following information:
\begin{itemize}
\item Server hostname/IP address
\item Network port
\item User name
\item Password
\item Virtual host name
\item Exchange name (required for publishing only)
\item Queue name (required for subscribing only)
\item Routing key (optional)
\end{itemize}

We initially wanted to be able to specify all the parameters in the port definitions like this:

\begin{lstlisting}
  outputPort someOutputPort {
    Location: ``amqp://amqpServer:port/vhost''
    Interfaces: SomeOutputInterface
    Protocol: sodep
    Username: user
    Password: pass
    Exchange: exch
    RoutingKey: rtKey
  }
  inputPort someInputPort {
    Location: ``amqp://amqpServer:port/vhost''
    Interfaces: SomeInputInterface
    Protocol: sodep
    Username: user
    Password: pass
    Queue: queue
    RoutingKey: rtKey
  }
\end{lstlisting}
However, this required that we changed core Jolie code. We did not want to have to change Jolie to be able to implement our extension. Our extension should be able to be copied to peoples Jolie installations, and work out-of-the-box. This meant that we ended up implementing the transfer of all the parameters using the location url:

\begin{lstlisting}
  outputPort SomeOutputPort {
    Location: ``amqp://user:pass@amqpServer:port/vhost?exchange=exch&routingkey=rtkey''
    Interfaces: SomeOutputInterface
    Protocol: sodep
  }
  inputPort SomeInputPort {
    Location: ``amqp://user:pass@amqpServer:port/vhost?queue=queue&routingkey=rtkey''
    Interfaces: SomeInputInterface
    Protocol: sodep
  }
\end{lstlisting}
Jolie has a method for extracting the query-part of an url, and we ended up writing a method for splitting the arguments into a map, for easier access.

According to the Jolie language, we need to implement the AmqpCommChannel and AmqpCommChannelFactory classes.
\subsubsection{AmqpCommChannel \& AmqpCommChannelFactory}
The following methods needs to be partially implemented to be able to publish to an exchange:

\noindent\textbf{AmqpCommChannel}\\
The constructor for our initial implementation was responsible for establishing a connection to the location specified on the port, and save it in the instance. We establish a connection this way:
\begin{lstlisting}
  // Connect to the AMQP server.
  String schemeAndPath = location.toString().split(``\\?'')[0];
  ConnectionFactory factory = new ConnectionFactory();
  factory.setUri(schemeAndPath);
  conn = factory.newConnection();

  // Create the channel.
  chan = conn.createChannel();
\end{lstlisting}
When publishing, subscribing or otherwise interacting with the server, the `chan' (com.rabbitmq.client.Channel) object is used.

\noindent\textbf{sendImpl}\\
When implementing the sendImpl method, we faced many difficulties. We took an incremental approach where we took small steps towards the final solution. To begin with, we implemented sending a simple hardcoded value to the RabbitMQ server. In the web-based administrative interface of the RabbitMQ server, we could see the messages on a specific queue, and could therefore verify that our message reached the queue we intended. Our first implementation ended up looking something like this:
\begin{lstlisting}
  chan.basicPublish(exchName, routingKey, null, ``hey''.getBytes())
\end{lstlisting}

When we got that working we tried to use the protocol defined. This meant adding the following lines:
\begin{lstlisting}
  // Make protocol give us the bytes to send.
  ByteArrayOutputStream ostream = new ByteArrayOutputStream();
  protocol().send(ostream, message, null);
\end{lstlisting}
And sending the byte array from the output stream.

In the early stages of our implementation we used the Jolie protocol SODEP (Simple Operation Data Exchange Protocol)\cite{SODEP}, but the messages it created held attributes only used for returning calls. Instead we created a protocol for sending simple values instead of the SODEP object which contains much more meta data. We simply copied the SODEP implementation and removed most of its features to create the SVDEP (Simple Value Data Exchange Protocol) protocol. This protocol can be used with for example AMQP to send values to clients that does not support SODEP. See \ref{subsec:SVDEP} on page \pageref{subsec:SVDEP} for the explanation of SVDEP protocol. The best solution would be to implement the AMQP type system in the SVDEP protocol (and rename the protocol), to fully support the capabilities of AMQP.

\noindent\textbf{recvImpl}\\
Because we are implementing the OneWay type, recvImpl is not interesting. We did however face some difficulties, as Jolie invokes the method regardless of whether or not there is a response to handle. To test our implementation we started by having Java print a line to the console, to indicate that the method has been called.

\noindent\textbf{closeImpl}\\
The implementation of closeImpl should just close the connection we have established in the constructor.

\noindent\textbf{The factory}\\
The factory class just creates an instance of the AmqpCommChannel class and returns it.

\subsubsection{AmqpCommListener \& AmqpCommListenerFactory}
One thing is being able to publish to an exchange, to fully implement OneWay we have to be able to subscribe to a queue as well, when acting as the other end. This means implementing the AmqpCommListener and its factory class AmqpCommListenerFactory

The following methods need to be partially implemented to be able to subscribe to a queue:

\noindent\textbf{AmqpCommListener}\\
The constructor is responsible for creating a connection to the AMQP server, and for subscribing to the queue.

We started by trying to implement a blocking call for use in the run method, but the AMQP client does not have such a method. Instead, it has the option to setup an instance of the Consumer object for handling when an item from a queue has been sent to us. This object has one method that needs to be implemented; handleDelivery. Fortunately, RabbitMQs client library has created a default consumer that we can create an anonymous class of. The handleDelivery method is given the body of the message, along with some properties.

To start with, we just printed the message we received from the AMQP server. Later we changed the implementation to create an instance of the AmqpCommChannel class, and use it for handling the delivery. Jolie has a method for handling a receive using a CommChannel that which used.

Jolie will call the recvImpl method of the CommChannel, and we had to modify this to be able to handle receiving a message. The recvImpl method does not take any arguments, so we need to save the data to process on the CommChannel object. We implemented the use of the protocol specified like this:
\begin{lstlisting}
  ByteArrayOutputStream ostream = new ByteArrayOutputStream();
  return protocol().recv(new ByteArrayInputStream(dataToProcess.body), ostream);
\end{lstlisting}

\noindent\textbf{run}\\
As the library creates a new thread for handling a delivery, the run method does not need to do anything. The entire run method just consists of an infinite loop with a call to .wait() on an object. This means that this thread does nothing, and consumes a minimum amount of resources.

\noindent\textbf{shutdown}\\
Implementing shutdown just meant closing our connection to the AMQP server.

We tested everything every step of the way, and found that our implementation (as expected) did only work for OneWay service-calls. We needed to support using RequestResponse calls as well to respect the Jolie guarantees. This meant looking into RPC calls using AMQP. AMQP is not supposed to be used as an RPC-style protocol, but with a bit of ingenuity it is possible.

\subsection{RequestResponse}
The second part of our implementation was focused on getting RPC-style calls working for the AMQP implementation. This meant modifying the AmqpCommChannel class we created ealier. For understanding how RPC works in AMQP, see \ref{subsubsec:AMQPRPC} on page \pageref{subsubsec:AMQPRPC}.

The RabbitMQ AMQP client library has support for RPC style communication through the RpcClient and RpcServer classes. RpcServer requires that you create a subclass, and does not work with how we have setup our Jolie extension. We ended up not using the RpcServer class, just the RpcClient.

To allow for RPC calls we have to modify the following methods:

\noindent\textbf{sendImpl}\\
First of all, handling RPC is different that regular OneWay, so we needed a way of determining whether or not the call we are currently handling is of the RequestResponse type. Jolie has the ability to give us the definition of the method through its interface definition. Through this we can determine the type.

If we are handling an RPC call we first create a new instance of the RpcClient class. It needs the AMQP Channel, the exchange to publish to, and the routing key. Because we want RPC calls to publish directly to queues, we set the exchange to be blank, and the routing key to be the name of the queue.

After creating this instance, we use the primitiveCall method to make a call to another method on the other side. This just gets the bytes to send to the queue.

The entire RPC client implementation ended up looking like this:
\begin{lstlisting}
  RpcClient rpc = new RpcClient(chan, ``'', queueName);
  dataToProcess = new AmqpMessage(null, rpc.primitiveCall(ostream.toByteArray(), null);
\end{lstlisting}

\noindent\textbf{recvImpl}\\
When we are at the other end, and needs to send back a response, we use the replyTo property of the properties that follows the message. We simply publish our response to that queue.


\subsection{SVDEP protocol}
\label{subsec:SVDEP} % For referencing this part elsewhere
The SVDEP protocol is used for when we wish to only transfer a value, and no metadata for it. This protocol could be extended to convert from the AMQP type system to the Jolie type system and vice versa. For now, we use it to be able to publish simple messages to an exchange, and for getting simple values from a queue.

We have basically just copied the SODEP implementation, changed the method writeMessage to only add the actual value to the stream, and removed the now unused methods.

\subsection{Concurrency problems and solutions}
When implementing both OneWay and RequestResponse, we experienced many periodic problems which turned out to be concurrency problems. This section will explain some of them, and their solutions.

\noindent\textbf{OneWay receiving and creating the CommChannel object}\\
At first, because we did not want to create several connections and channels to a single AMQP location we did not create a new object for each message we received from our subscription. Instead, we created one instance of the AmqpCommChannel object and saved that in our AmqpListener instance. This resulted in concurrency problems, as all the threads used the same instance, and race conditions sometimes occurred.

We spent a lot of time thinking about how to fix this, and ended up creating a handler class for AMQP connections. This class acts like a singleton having one connection for each location. The connections are stored by their location, and when you need a connection you give it the location. If there is no such connection, the handler will create it, and open a channel.

By using this new connection handler, the creation of AmqpCommChannel will be instant, and not be dependent on network. Whenever we need to do something with the server (publish, subscribe, RPC primitiveCall) we would use the handler to get the connection.

\noindent\textbf{Matching a response with the requester}\\
When a request comes in the extension (simply put) passes the request on to Jolie. When the user written program has found an appropriate response, this is passed to Jolie and in turn to our extension. It was also difficult to identify which response matched which request, when the requests came tumbling in, but we did so by using the AMQP message property 'correlationId'.

This problem was even better solved with the same solution mentioned above. With the implementation of the solution where there is a single connection per location we create an instance of AmqpCommChannel per request. When the user written program returns the respond, Jolie passes it to the specific CommChannel instance and thereby the problem is resolved in a better manner.

We further improved the efficiency of the extension by moving the field for the RpcClient out of the AmqpCommChannel and thereby having just a single RpcClient instead of one per request. The RpcClient then relies on the correlationId to match the corresponding requester to a response.
\newpage